Probability of causal contact between interstellar civilizations through MonteCarlo simulations

### 2019-08-28 manuscript.tex
biblio...

A common approach to discuss the possible number of extraterrestrial
intelligent civilizations (ETI) has been the use of the Drake equation
\citep{gleiser_drake_2010, prantzos_joint_2013, haqq-misra_drake_2017}.
%
However, the uncertainties in the factors make it less prone to a
formal application to define searching strategies or to compute
actual numbers, but is used to organize the discussion instead.
%
Some modifications to the original idea of the equation have tried to
imprint a stochastic nature, or to propose a probabilistic approach,
or to consider the temporal structure which is missing in the
equation.
%
Temporal aspects of the distribution of communicating civilizations
and their contacts have also been explored by several authors
\citep{fogg_temporal_1987,forgan_spatiotemporal_2011, balbi_impact_2018},
%The Impact of the Temporal Distribution of Communicating Civilizations on Their Detectability
%Temporal dispersion of the emergence of intelligence: an inter-arrival time analysis
%The spatiotemporal aspects of SETI
%
as well as efforts on considering the stochastic nature of the Drake equation
\citep{glade_stochastic_2011}.
%
The simulation approach has also benn considered
\citep{forgan_evaluating_2015, vukotic_grandeur_2016,
murante_simulating_2015, forgan_numerical_2009, forgan_galactic_2017}
%New numerical determination of habitability in the Galaxy: the SETI connection
%
Since the factors in the Drake equation are uncertain, we propose to
avoid the frequentist approach of this equation
and to explore a parameter space, where instead of computing a final
number, we provide a statistical distribution that gives conditional
probabilities.
%
The only fact that can be stated with certainty is that for the number
of years SETI projects have been working we have not received any
signal within the conditions stablished by SETI.
%
This, by some astonishing result, has lead to the Fermi paradox, which
states an apparent contradiction between the high estimation of the
abundance of life in the Galaxy and the lack of their detection
\citep{anchordoqui_isthere_2017}.

@@

per unit time.

@@

There is no a temporal window for the model, i.e., the simulation
starts when the process is already stable, and ends before any
galactic effect could modify or affect of fixed variables.
%
The probability for the rise of a CETI is homogeneous over the GHZ.
%
Although the Galaxy has a well known spiral structure, the assumed
sparcity of civilizations makes this assumption very reasonably.
%
Otherwise, if the distribution of civilizations is not sparce, it
could be the case that the spiral arms would host most of the
civilizations, and contacts are frecuent between closely located
civilizations.
%
In such a case we could expect that the interpretation of our results
could be considered as pesimist for the stablishment of a contact with
our planet.
%
We also limit the posibilities of life or other types of civilizations
to the hypothesis usually stated for the definition of the GHZ.
%
This means that we set aside possibel civilizations that could survive
in severe conditions or unstable systems, which would prevent the
appeareance of life as we know it.

@@

In our simulations we also assume the simplest situations for the
growth of ETIs.
%
For example, we discard the possibility of stellar colonization
(Walters 1980) and assume that the communication aim is performed to
all directions with the same power and the same probability, i.e.,
strongly assume isotropic communication in all cases.
%
More detailed simulations could be produced considering different 
efficiency of communication or detection methods.
%
Also, a mixed paranoid/pronoid population of intelligent civilizations
could be implemented easily.
%
However, in this case we also would complicate the experimental setup,
make the results less evident.
%
The results are independent of the nature of life (organic or
artificial).
%
The lifetime of a civilization can be caused by autodestruction or by
external factors.
%}}}

@@

The power law and exponential statistical distributions are among the
most common patterns found in natural phenomena.
%
For example, the distribution of the frecuency of words in many
languajes is known to follow a Zipf law (which is a power law).
%
Zipf law also describes population ranks of cities in various
countries, corporation sizes, income rankings, ranks of number of
people watching the same TV channel (Zipf's law entry on Wikipedia).
%
The magnitudes of earthquakes, hurricanes, volcanic eruptions and
floods; the sizes of meteorites or the losses caused by business
interruptions from accidents, are also well described by power laws
(Sornette 2006).
%
The power law behaviour has been observed in a variety of systems,
including for example stock market fluctuations, sizes of computer
files and word frequency in languages \citep{mitzenmacher_brief_2004,
newman_power_2005, simkin_theory_2006}
%
Power laws have also been widely used in biological sciences, e.g., in
analysing connectivity patterns in metabolic networks
\citep{jeong_large_2000} and in the number of species observed per
unit area in ecology \citep{martin_origin_2006, frank_common_2009}.

@@

%\subsection{Exploration of the hypothesis space}

@@

In the Fig.~\ref{F_waiting_time} we show the fraction of the CETIs
that have made at least one contact as a funtion of the elapsed time
since the awakening, given that they make at least one contact on
their lifetime, for different models.
%
From a frequentist approach, the normalized cumulative counts are
related to the estimation of the probability of observing a given time
with no success.
%             
Both panels show the histograms of the waiting times for the first
contact, for all the CETIs that make at least one contact along their
lifetime.
%
We show separately the histograms for several values of $\tau_a$ and a
fixed value of $(\tau_s, D_{max})$, and for several values of $\tau_s$
and a fixed value of $(\tau_a, D_{max})$.
%
For different values of $\tau_a$, the shape of the distribution is
similar, and the main difference given by the different number of
contacts on the different models.
%
Conversely, for different values of $\tau_a$, the shape of the
distribution is similar, and the main difference given by the
different number of contacts on the different models. 
%
We show the 2D maps of the rate of CETIs that never make contact and
the rate of CETIs that make contact at the moment of the awakening,
inf the Figs.~\ref{F_never_contact} and \ref{F_C_at_A}, respectively,
as a function of $\tau_s$ and $\tau_s$.
%
In the Fig.~\ref{F_waiting_for_1C} we show the histograms of the mean
waiting times for the first contact, for several models.
%
Left panel shows the histograms for several values of $\tau_a$, and
$\tau_s$ in the range 17000-24000 yr.
%
Right panel shows the histograms for several values of $\tau_s$, and
$\tau_a$ in the range 68000-100000 yr.
%
The shape of the histograms does not change when $\tau_a$ is modified,
but significantly changes when $\tau_s$ is modified.

@@

Taking into account the caveat mentioned previously about the
trade--off between the simplicity and complexity of the experiment,
future work can be performed following this framework in order to
explore possible implications of the results for more detailed
configurations.
%
For example, the communication method (isotropic, colimated,
serendipitous) can affect the observables.
%
Since we analyze the causal contacts, effective contacts depend on the
efficiency of both the emitter and the receiver.
%
A correction by coverage ratio in the detection and by a targetting
ratio in the emission could be easily implemented in the simulation,
although the effect would be to reduce the probability of contact by
an efficiency ratio.
%
Thus, such implementation is not really necessary, since the values of
probabilities could modified by a constant correction factor equal to
the combined emission/reception efficiency.
%
If we consider a statistical distribution for this factor, then the
results could change, and this could be studied with the DE approach.
%
Also the nature of the message carrier, electromagnetic or another.
%
It is also possible to explore the effects of alignments or the use of
stars as sources or amplifiers.
%
%energetic markers Spiritual markers
%
As a future work, we will explore several improvements to the
simulation, and analyze the impact on the final results of these
changes.
%
They include a spatial distribution that actually resembles the shape
of the Galaxy, different distribution functions for the mean lifetime
of CETIs and different efficiencies for the sphere of the causal
region of each CETI, resembling different searching strategies.

@@

\begin{itemize}
   \item Dmax is different for different CETIs.  For example a power
      law where powerfull emmisions are rare and weak emssions are
      common.
   \item The probability of the rising of a new CETI vary within the
      GHZ.
   \item Spiritual factor.  Explore the role of the S factor: message
      contents, influence on the lifespan of a CETI that receives a
      message, mean lifespan of a CETI that emits a message.
\end{itemize}

@@

\nocite{barabasi_scale_2009}
\nocite{barlow_galactic_2012}
\nocite{bloetscher_using_2019}
\nocite{burchell_whither_2006}
\nocite{conway_three_2018}
\nocite{dayal_habitability_2016}
\nocite{enriquez_breakthrough_2017}
\nocite{forgan_exoplanet_2017}
\nocite{forgan_possibility_2013}
\nocite{forgan_collimated_2014}
\nocite{forgan_numerical_2010}
\nocite{funes_searching_2019}
\nocite{glade_stochastic_2012}
\nocite{gonzalez_habitable_2005}
\nocite{gonzalez_galactic_2001}
\nocite{grimaldi_signal_2017}
\nocite{haqq-misra_evolution_2019}
\nocite{harp_application_2018}
\nocite{hinkel_interdisciplinary_2019}
\nocite{hippke_interstellar_2017}
\nocite{horvat_calculating_2007}
\nocite{horvat_impact_2011}
\nocite{lampton_information_2013}
\nocite{lingam_relative_2019}
\nocite{loeb_relative_2016}
\nocite{loeb_eavesdropping_2006}
\nocite{maccone_SETI_2013}
\nocite{maccone_evolution_2014}
\nocite{maccone_mathematical_2011}
\nocite{maccone_statistical_2015}
\nocite{maccone_KLT_2010}
\nocite{maccone_statistical_2010}
\nocite{maccone_SETI_2011}
\nocite{morrison_extending_2015}
\nocite{morrison_extending_2015}
\nocite{newman_galactic_1981}
\nocite{peters_outer_2018}
\nocite{rahvar_cosmic_2016}
\nocite{seager_search_2018}
\nocite{shostak_what_2010}
\nocite{smith_broadcasting_2009}
\nocite{solomonides_probabilistic_2016}
\nocite{starling_virulence_2013}
\nocite{tarter_search_2001}
\nocite{tarter_advancing_2009}
\nocite{vukotic_astrobiological_2012}
\nocite{walters_interstellar_1980}
\nocite{wright_theGsearch_2015}
\nocite{anchordoqui_upper_2019}
\nocite{balb_spatiotemporal_2018}
\nocite{forgan_galactic_2016}
\nocite{gobat_evolution_2016}

>>>>

A common approach to discuss the possible number of extraterrestrial
intelligent civilizations (ETI) has been the use of the Drake equation
\citep{gleiser_drake_2010, prantzos_joint_2013, haqq-misra_drake_2017}.
%
The abundance of intelligent civilizations in the galaxy is a
longstanding question, which is often conceptualized as the problem of
the lack of received communication or the Fermi paradox
\citep{barlow_galactic_2012, Sotos_biotechnology_2019, 
forgan_galactic_2016}.
There are many proposals intended to resolve this paradox, which make
use of statistical \citep{solomonides_probabilistic_2016,
vanhouten_isthere_2017, horvat_calculating_2007, maccone_statistical_2015} or
stochastic approaches \citep{forgan_numerical_2009,
bloetscher_using_2019, glade_stochastic_2011, forgan_numerical_2010},
analytical interpretations of the Drake equation \citep{prantzos_joint_2013, smith_broadcasting_2009}
or more speculative proposals
\citep{barlow_galactic_2013, lampton_information_2013, conway_three_2018, forgan_galactic_2017}.
%
However, the uncertainties in the factors make it less prone to a
formal application to define searching strategies or to compute
numbers, but is used to organize the discussion instead
\citep{hinkel_interdisciplinary_2019}.
%
Some modifications to the original idea of the equation have tried to
imprint a stochastic nature, or to propose a probabilistic approach,
or to consider the temporal structure which is missing in the
equation.
%
Temporal aspects of the distribution of communicating civilizations
and their contacts have also been explored by several authors
\citep{fogg_temporal_1987, forgan_spatiotemporal_2011,
balbi_impact_2018, balb_spatiotemporal_2018, horvat_impact_2011},
%The Impact of the Temporal Distribution of Communicating Civilizations on Their Detectability
%Temporal dispersion of the emergence of intelligence: an inter-arrival time analysis
%The spatiotemporal aspects of SETI
%
as well as efforts on considering the stochastic nature of the Drake equation
\citep{glade_stochastic_2011}.
%
The simulation approach has also benn considered
\citep{forgan_evaluating_2015, vukotic_grandeur_2016,
murante_simulating_2015, forgan_numerical_2009, forgan_galactic_2017}
%New numerical determination of habitability in the Galaxy: the SETI connection
%
Since the factors in the Drake equation are uncertain, we propose to
avoid the frequentist approach of this equation and to explore a
parameter space, where instead of computing a final number, we provide
a statistical distribution that gives conditional probabilities.
%
The only fact that can be stated with certainty is that for the number
of years SETI projects have been working we have not received any
signal within the conditions stablished by SETI \citep{tarter_search_2001}.
%
This lack of detection has lead to a number of alternative proposals
for new SETI strategies \citep{forgan_exoplanet_2017,
balbi_impact_2018, loeb_eavesdropping_2006, maccone_KLT_2010,
tarter_advancing_2009, enriquez_breakthrough_2017, loeb_relative_2016,
maccone_SETI_2011,  lingam_relative_2019, wright_theGsearch_2015,
maccone_SETI_2013, maccone_lognormals_2014, harp_application_2018,
forgan_possibility_2013, forgan_galactic_2017, funes_searching_2019}

This, by some astonishing result, has lead to the Fermi paradox, which
states an apparent contradiction between the high estimation of the
abundance of life in the Galaxy and the lack of their detection
\citep{vanhouten_isthere_2017}.

@@

per unit time \citep{maccone_evolution_2014,Sotos_biotechnology_2019}.

@@

There is no a temporal window for the model, i.e., the simulation
starts when the process is already stable, and ends before any
galactic effect could modify or affect of fixed variables.
%
The probability for the rise of a CETI is homogeneous over the GHZ.
%
Although the Galaxy has a well known spiral structure, the assumed
sparcity of civilizations makes this assumption very reasonably.
%
Otherwise, if the distribution of civilizations is not sparce, it
could be the case that the spiral arms would host most of the
civilizations, and contacts are frecuent between closely located
civilizations.
%
In such a case we could expect that the interpretation of our results
could be considered as pesimist for the stablishment of a contact with
our planet.
%
We also limit the posibilities of life or other types of civilizations
to the hypothesis usually stated for the definition of the GHZ
\citep{dayal_habitability_2016,gonzalez_galactic_2001,lineweaver_galactic_2004,gonzalez_habitable_2005,morrison_extending_2015,haqq-misra_evolution_2019,rahvar_cosmic_2016,gobat_evolution_2016,
rahvar_cosmic_2016}.
%
\citet{vukotic_astrobiological_2012} propose probabilistic cellular
automata model to explain the astrobiological history of the Milky
Way.
%
This means that we set aside possibel civilizations that could survive
in severe conditions or unstable systems, which would prevent the
appeareance of life as we know it.

@@

In our simulations we also assume the simplest situations for the
growth of ETIs.
%
For example, we discard the possibility of stellar colonization
\citep[e.g.][]{newman_galactic_1981, walters_interstellar_1980,
starling_virulence_2013, barlow_galactic_2012, jeong_large_2000, maccone_mathematical_2011}
and assume that the communication aim is performed to
all directions with the same power and the same probability, i.e.,
strongly assume isotropic communication in all cases.
%
More detailed simulations could be produced considering different 
efficiency of communication or detection methods.
%
Also, a mixed paranoid/pronoid population of intelligent civilizations
could be implemented easily.
%
However, in this case we also would complicate the experimental setup,
make the results less evident.
%
The results are independent of the nature of life (organic or
artificial).
%
The lifetime of a civilization can be caused by autodestruction or by
external factors.
%}}}

@@

The power law and exponential statistical distributions are among the
most common patterns found in natural phenomena.
%
For example, the distribution of the frecuency of words in many
languajes is known to follow a Zipf law (which is a power law).
%
Zipf law also describes population ranks of cities in various
countries, corporation sizes, income rankings, ranks of number of
people watching the same TV channel (Zipf's law entry on Wikipedia).
%
The magnitudes of earthquakes, hurricanes, volcanic eruptions and
floods; the sizes of meteorites or the losses caused by business
interruptions from accidents, are also well described by power laws
(Sornette 2006).
%
The power law behaviour has been observed in a variety of systems,
including for example stock market fluctuations, sizes of computer
files and word frequency in languages \citep{mitzenmacher_brief_2004,
newman_power_2005, simkin_theory_2006}
%
Power laws have also been widely used in biological sciences, e.g., in
analysing connectivity patterns in metabolic networks
\citep{jeong_large_2000} and in the number of species observed per
unit area in ecology \citep{martin_origin_2006, frank_common_2009}.
% 
More examples can be found in the literature
\citep{martin_origin_2006, maccone_KLT_2010, barabasi_scale_2009,
maccone_evolution_2014, maccone_lognormals_2014,
benguigui_classificacion_2016}

@@

@@

In the Fig.~\ref{F_waiting_for_1C} we show the fraction of the CETIs
that have made at least one contact as a funtion of the elapsed time
since the awakening, given that they make at least one contact on
their lifetime, for different models.
%
From a frequentist approach, the normalized cumulative counts are
related to the estimation of the probability of observing a given time
with no success.
%             
Both panels show the histograms of the waiting times for the first
contact, for all the CETIs that make at least one contact along their
lifetime.
%
We show separately the histograms for several values of $\tau_a$ and a
fixed value of $(\tau_s, D_{max})$, and for several values of $\tau_s$
and a fixed value of $(\tau_a, D_{max})$.
%
For different values of $\tau_a$, the shape of the distribution is
similar, and the main difference given by the different number of
contacts on the different models.
%
Conversely, for different values of $\tau_a$, the shape of the
distribution is similar, and the main difference given by the
different number of contacts on the different models. 
%
We show the 2D maps of the rate of CETIs that never make contact and
the rate of CETIs that make contact at the moment of the awakening,
inf the Figs.~\ref{F_never_contact} and \ref{F_C_at_A}, respectively,
as a function of $\tau_s$ and $\tau_s$.
%
In the Fig.~\ref{F_waiting_for_1C} we show the histograms of the mean
waiting times for the first contact, for several models.
%
Left panel shows the histograms for several values of $\tau_a$, and
$\tau_s$ in the range 17000-24000 yr.
%
Right panel shows the histograms for several values of $\tau_s$, and
$\tau_a$ in the range 68000-100000 yr.
%
The shape of the histograms does not change when $\tau_a$ is modified,
but significantly changes when $\tau_s$ is modified.

@@

Taking into account the caveat mentioned previously about the
trade--off between the simplicity and complexity of the experiment,
future work can be performed following this framework in order to
explore possible implications of the results for more detailed
configurations.
%
For example, the communication method (isotropic, colimated,
serendipitous) can affect the observables.
%
Since we analyze the causal contacts, effective contacts depend on the
efficiency of both the emitter and the receiver.
%
A correction by coverage ratio in the detection and by a targetting
ratio in the emission could be easily implemented in the simulation,
although the effect would be to reduce the probability of contact by
an efficiency ratio.
%
Thus, such implementation is not really necessary, since the values of
probabilities could modified by a constant correction factor equal to
the combined emission/reception efficiency
\citep{smith_broadcasting_2009, anchordoqui_upper_2019, forgan_collimated_2014}.
%
If we consider a statistical distribution for this factor, then the
results could change, and this could be studied with the DE approach.
%
Also the nature of the message carrier, electromagnetic or another.
%
It is also possible to explore the effects of alignments or the use of
stars as sources or amplifiers.
%
%energetic markers
%Spiritual markers
%
As a future work, we will explore several improvements to the
simulation, and analyze the impact on the final results of these
changes.
%
They include a spatial distribution that actually resembles the shape
of the Galaxy, different distribution functions for the mean lifetime
of CETIs and different efficiencies for the sphere of the causal
region of each CETI, resembling different searching strategies
\citep{hippke_interstellar_2017}.

@@

\begin{itemize}
   \item Dmax is different for different CETIs.  For example a power
      law where powerfull emmisions are rare and weak emssions are
      common.
   \item The probability of the rising of a new CETI vary within the
      GHZ.
   \item Correction by coverage ratio in the detection and by
      targetting ratio in the emission.
   \item Spiritual factor.  Explore the role of the S factor: message
      contents, influence on the lifespan of a CETI that receives a
      message, mean lifespan of a CETI that emits a message.
\end{itemize}

@@
